Practical Optional Types for Clojure

### 2015-02-23 formal-model.tex
java interop section 2

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields. We model
these operations in a restricted setting without class inheritance,
overloading or Java Generics.

@@

Clojure supports \emph{type hints}, an optional type system to improve performance
and disambiguate Java calls.
Programmers communicate to the Clojure compiler via type hints
to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
and the programmer is free to add them where they like.

>>>>

We model
Java interoperability in a restricted setting without class inheritance,
overloading or Java Generics.

@@

### 2015-02-23 overview.tex
java interop section 2

Here we include the full namespace, which we omit in the following
examples. Going down the then branch in this example means that we know
\clj{x} is a number, which is automatically updated with occurrence typing.

In Clojure, exceptional control flow is almost as common, often in the form of
runtime assertions.

@@

Our main insight is that occurrence typing already provides
valuable information about the left argument,
namely if it returns a value, \clj{x} is a \clj{Number}.
Since the right argument is evaluated after the left, we can assume
\clj{x} is a \clj{Number} when evaluating \clj{(inc x)}, verifying
the programmer's attempt to avoid null-pointer exceptions without further annotations.

@@

is the type of a map with at least an \clj{:a} entry mapped to a number,
also written \clj{'{:a Num}}. If \clj{:b} is known to be an absent entry
then

@@

We use \clj{defalias} to abbreviate HMap types.

@@

The result type is updated as expected to exclude \clj{nil}.
Testing for \clj{nl} is more complicated, since it is the sentinel
lookup failure indicator but may also appear as an entry value.
If we test for \clj{nil} in an \clj{:optional} entry, the best we learn
is that either the entry is missing or contains \clj{nil}.

@@

\smallsection{Mandatory Entries}

\smallsection{Known Absent Entries}

\smallsection{Unknown Entries}

@@

Java

\begin{listing}
\inputminted[firstline=5,lastline=14]{clojure}{code/demo/src/demo/hmap.clj}
\caption{Java Interoperability}
\end{listing}

>>>>

Here we include the full \clj{ns} form, which we omit in the following
examples. Going down the then branch in this example means that we know
\clj{x} is a number, which is automatically updated with occurrence typing.

In Clojure, exceptional control flow is also common, often in the form of
runtime assertions.

@@

Our main insight is that occurrence typing already provides
valuable information about the left argument of \doliteral{} expressions,
namely if it returns a value, \clj{x} is a \clj{Number}.
Since the right argument is evaluated after the left, we can assume
\clj{x} is a \clj{Number} when evaluating \clj{(inc x)}, verifying
the programmer's attempt to avoid null-pointer exceptions without further annotations.

@@

is the type of a map with at least an \clj{:a} entry mapped to a number,
also written \clj{'{:a Num}}. If \clj{:b} is also known to be an absent entry
then

@@

\clj{defalias} defines a custom type abbreviation, often for unions of HMaps.

@@

The result type is updated as expected to exclude \clj{nil}.
Testing for \clj{nil} is more complicated, since it is the sentinel
lookup failure indicator but may also appear as an entry value.
If we test for \clj{nil} in an \clj{:optional} entry, the best we learn
is that either the entry is missing or contains \clj{nil}.

@@

@@

Clojure supports interoperability with Java, including the ability to
call constructors, methods and access fields.

Clojure supports \emph{type hints}, an optional type system to improve performance
and disambiguate Java calls.
Programmers communicate to the Clojure compiler via type hints
to resolve Java methods and prevent boxing in loops. Tags are propagated bidirectionally
and the programmer is free to add them where they like.

Here we use \clj{ann} to annotate static types for the function, and the type hint
\clj{^File} directs the Clojure compiler to assume \clj{f} is a \clj{File}
when resolving Java calls at compile time.

\begin{listing}
  \begin{minted}{clojure}
(ann parent [(U nil File) -> (U nil Str)]
(defn parent [^File f]
  (when f
    (.getParent f)))
  \end{minted}
\caption{Java Interoperability}
\end{listing}

While \clj{f} has the type hint \clj{File}, it also has the static type \clj{(U nil File)},
and Typed Clojure disallows invoking \clj{nil} even if the compiler
has statically resolved a method.
Typed Clojure also conservatively assumes fields and method returns are nullable,
and guarantees null-pointer exceptions cannot happen in typed code.
